---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Installation

```{r, eval=FALSE}
devtools::install_github('mcbg/AptamerTools')
```

# AptamerTools

Contrived example since the data is already normalized using ANML. 

```{r}
library(magrittr)
library(data.table)
library(knitr)
library(AptamerTools)

temp <- tempfile()
url <- 'https://github.com/SomaLogic/SomaLogic-Data/raw/master/example_data.adat'
download.file(url, destfile = temp)

ds <- SomaDataIO::read_adat(temp)

# extract metadata
meta <- ds %>% attributes() %>% .$Col.Meta %>% data.table()
meta[, variable := gsub('-', '.', SeqId) %>% paste0('seq.', .)]

# convert to data.table
setDT(ds)
```

Somalogic provides data in a wide format. Here we transform to a long format.

```{r}
seq_cols <- names(ds) %>% { .[grepl('^seq', .)] }

ds.long <- ds %>% 
  melt(measure.vars = seq_cols) %>% 
  merge(meta, by = 'variable')
```

# Normalization example

The following code derives the scale factors of each step in the pipeline and the normalized data.
The variable `value_HIPCA` contains the fully normalized measurement. Note that the ANML step isn't recommended by the author.

```{r, results='hide'}
# derive references
ds.long[SampleType %in% c('Sample', 'QC'), ref.sample := median(value), .(variable)]
ds.long[SampleType %in% c('Sample'), mad := mad(value), .(variable)]
ds.long[, ref.hyb := median(value), .(variable)]
ds.long[SampleType == 'Calibrator', ref.cal := median(value), .(variable)]

# apply normalization
normalization_pipeline(
  ds.long,
  'value',
  steps = list(hyb_norm, intraplate_norm, plate_norm, calibration_norm, anml_norm),
  step_names = c('H', 'HI', 'HIP', 'HIPC', 'HIPCA'),
  refs = list('ref.hyb', derive_intraplate_ref, 'ref.cal', 'ref.cal', 'ref.sample')
)
```
```{r, results='asis'}
ds.long[, .(SampleId, variable, value_HIPCA)] %>% 
  head() %>% 
  kable()
```

# Analysis

In the paper we recommend either robust regression or non-parametric methods to analyse SomaScan data.

## Robust regression using MM-estimation

Uses the package `RobStatTM`.

```{r, results='asis'}
library(RobStatTM)

analysis.MM.estimation <- ds.long[, {
    ml <- lmrobdetMM(log2(value_HIPCA) ~ Sex, data = .SD)
    coef <- ml %>% summary() %>% .$coefficients
    pvalue <- coef %>% .[2, 4]
    logFC <- coef %>% .[2, 1]
    converged <- ml$converged
    data.table(pvalue, logFC, FC = 2^logFC, converged)
}, .(variable)]

head(analysis.MM.estimation) %>% kable()
```

## Robust regression using M-estimation
Uses the package `RobStatTM`.
```{r, results='asis'}
# settings for regression
control <- lmrobM.control(bb = 0.5, efficiency = 0.95, family = 'bisquare', mscale_maxit = 1000, max.it = 1000)

analysis.M.estimation <- ds.long[, {
    ml <- lmrobM(log2(value_HIPCA) ~ Sex, data = .SD, control = control)
    coef <- ml %>% summary() %>% .$coefficients
    pvalue <- coef %>% .[2, 4]
    logFC <- coef %>% .[2, 1]
    converged <- ml$converged
    data.table(pvalue, logFC, FC = 2^logFC, converged)
}, .(variable)]

head(analysis.M.estimation) %>% kable()
```


## Kruskal-Wallis (non-parametric)
```{r, results='asis'}
# settings for regression
control <- lmrobM.control(bb = 0.5, efficiency = 0.95, family = 'bisquare', mscale_maxit = 1000, max.it = 1000)

analysis.kruskal <- ds.long[, {
    ml <- kruskal.test(value_HIPCA ~ Sex, data = .SD)
    pvalue <- ml$p.value
    data.table(pvalue)
}, .(variable)]

head(analysis.kruskal) %>% kable()
```
